Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance

### 2022-06-14 text/Manuscript.Rmd
MS: added abstract, added comments with Markdown syntax

---
title: "Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance in zooplankton"
author: "Kei Jokura, Nobuo Ueda, Martin Gühmann, Gáspár Jékely"
date: "*Living Systems Institute, University of Exeter, Stocker road, Exeter, EX4 4QD, UK*"
output:
  html_document: 
    theme: lumen
  pdf_document: 
      latex_engine: xelatex
      template: "default"
geometry: left=0.8in,right=0.8in,top=0.8in,bottom=0.5in
mainfont: Arial
documentclass: article
classoption:
  - onecolumn
linestretch: 1.2
fontsize: 10pt
bibliography: references.bib
csl: elife.csl
---

@@

Nitric oxide is... Here we show...

@@

The ability to perceive changes in the environment and adapt behaviour accordingly is essential for the survival of all living organisms. Avoidance from UV radiation, which damages DNA essential for survival, is proposed to be one of the behaviours that existed before the abundant development of Earth's oxygen to block UV radiation in the atmosphere (Pittendrigh). In the ocean, almost all zooplankton have a pronounced rhythmic behaviour called Diel vertical migration (DVM), which consists of upwards towards the surface at dusk and descend back to deeper water before dawn. Avoidance behaviour to escape harmful UV radiation during the day is closely linked to the control of vertical migration in the ocean. In metamorphosis, a life history strategy widely found in metazoans, downward swimming towards the seafloor is known to be associated with larval settlement and the initiation of metamorphosis, a dramatic body change from the larval stage, and UV-avoidance behaviour allows dispersal and the acquisition of different ecological niches. However, little is known about the molecular neural circuit mechanisms in UV avoidance behaviour in many animals, including zooplankton. The cPRCs express the UV-absorbing c-opsin1 and mediate UV avoidance in the larvae [@veraszto2018]. In adult worms, UV light and c-opsin1 modulate locomotion and maturation [@Veedin2021]. In neural circuits in general light response, the activation of various receptors by excitatory and inhibitory neurotransmitters such as glutamate, GABA and dopamine is known in various species. In addition to this, it has long been known that in nitric oxide (NO), as a neuromodulatory molecule, it is also important as a modulator of visual activity. NO is a small free-radical molecule with diffusible and changeable characteristics, and thus works as one of the neural signalling molecules. The function of NO in the neural circuitry of vision has been well studied in vertebrates. In the mammalian retina, NO is an essential signalling molecule in all retinal cells. NO produced by nitric oxide synthase (NOS) in amacrine cells, one of the retinal cells, activates soluble guanylate cyclase (sGC) expressed in cone bipolar cells, which produces cGMP and regulates downstream signalling. On the other hand, the role of NO in zooplankton has been reported in many animals, including sponges, cnidarians, molluscs, arthropods, annelids, echinoderms and ascidians, as a molecule that controls settlement and metamorphosis initiation. We therefore hypothesise that NO-mediated neural signalling may be involved in the photoreceptor circuits that control UV avoidance behaviour. To test this hypothesis we used the annelid *Platynereis dumerilii*, which conserved vertebrate-type ciliary photoreceptors (cPRCs) [@Arendt2004]. The cPRC expresses c-opsin1, which absorbs UV radiation, is involved in melatonin production and regulates circadian swimming behaviour [@Tosches2014; @tsukamoto2017]. Furthermore, we previously reported abnormal UV avoidance behaviour in c-opsin1 knockout larvae [@Veraszto2018]. These suggest that Platynereis larvae are a highly suitable model for elucidating the neural circuit mechanisms of NO-induced UV avoidance behaviour, and the accessibility of genome editing, behavioural assays, calcium imaging and whole-body connectomics could be very useful for study at the molecular level[@ozpolat2021].

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data. Phylogenetic analysis of NOS proteins indicate that the *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences (#Kei - need to add tree). To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis. In two-day-old larvae, we detected *NOS* expression in four cells in the apical organ region (Fig. 1A) (#Kei - 3d HCR and transgene?). The same four cells cells were labelled with a reporter construct driving palmitoylated-tdTomato by a 12 kbp upstream genomic region of the *NOS* gene. This reporter also revealed the axonal projections of the NOS-expressing neurons. The position and morphology of the four *NOS*-expressing cells allowed us to identify the same cells in the three-day-old whole-body volume EM data [@williams2017; @veraszto2020] as four interneurons (INNOS)(Figure 1D-E). The INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW interneurons, which are also cPRC targets (Fig. 1F).

@@

@@

The specific expression of NOS in interneurons directly postsynaptic to the cPRCs suggests that NO signalling is involved in the UV response. To test if NO is produced during UV/violet responses, we injected zygotes with the fluorescent NO-reporter DAF-FM and imaged two-day-old larvae during the local stimulation of the cPRC cilia by 405 nm light. Following light stimulation, we detected an increase in DAF-FM fluorescence in the region of the INNOS cells. (#Kei - are there time course data that could be shown?). We next analysed the dynamics of the cPRC circuit following light stimulation. As we have shown previously [@Veraszto2018], a 30-sec illumination of the cPRC ciliary region lead to a decrease in cPRC calcium levels, likely due to the hyperpolarisation of the cPRCs. After 30 min, calcium levels in cPRCs were raising again, reaching higher levels than at the start of the stimulus - a response that may involve depolarisation. This activation phase ocurres after the end of the stimulation and may be due to a slower neuroendocrine feedback [@Veraszto2018]. We also quantified in more detail the responses of the INNOS and INRGW interneurons under a UV/violet stimulus to the cPRCs. To unambiguously identify the cells from which we recorded calcium signals, we developed an on-slide staining method to directly fix, immunostain and image the same larva that we imaged previously. We used the cell-specific neuropeptide markers anti-RYamide (INNOS) and anti-RGWa (INRGW) to mark the interneurons. Based on the position of the nuclei, we could correlate live and fixed samples at a single-cell precision. INNOS cells showed increase calcium signals during the inhibition phase of the cPRCs. In contrast, INRGW cells were only active during the second activation phase of cPRCs.\
To determine whether NO is involved in the observed circuit dynamics, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method. We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both leading to an early stop codon and thus likely representing null alleles (Figure 3A). We could establish a homozygous line for both mutations indicating that NOS is not an essential gene. Calcium imaging in *NOS*-mutant larvae revealed perturbed cPRC responses and circuit activation. In cPRCs, we could detect the inhibition phase but this was not followed by activation and calcium levels remained low. We did not detect a difference in the response of the INNOS cells between wild-type and *NOS*-mutant larvae. In contrast, INRGW cells did not activate in *NOS*-mutant larvae, suggesting that they will not initiate a circuit output (to Ser-h1 cells and the ciliary band).

**Nitric oxide signalling mediates UV-avoidance behaviour** Our calcium imaging data suggest that *NOS*-mutant larvae will have defective UV-avoidance behaviour. To test this, we recorded the trajectories of freely swimming larvae in vertical columns, illuminated from two lateral sides with xy nm violet light. As previously shown, wild-type larvae swim downward within 30 sec following UV/violet light stimulation. In contrast, homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response (Figure 3B). This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae.

![**Figure 2. Larvae with inhibited NOS have shorter ciliary arrest times and a delayed response to UV-avoidance.** (A) Percentage of time in which cilia are closed in DMSO-treated larvae (control) vs. larvae exposed to NOS inhibitor. (B) Top: The domain structure of NOS. Bottom: Close-up region showing the genomic locus of NOS gene and the wild-type sequence (WT) targeted by CRISPR/Cas9. The generated mutants (NOSΔ11, NOSΔ23) are also shown. Pink indicates target sites. Yellow shows PAM sequences, red shows stop codons. (C and D) Vertical swimming in wild type and mutant larvae at 2-day-old (C) and 3-day-old (D) stimulated with blue (480 nm) and UV (395 nm) light from above. the data are shown in 30 s bins.](../figures/Fig2.png){width="100%"}

@@

Generally, NO can act via cyclic guanosine monophosphate (cGMP) as a second messenger. In this signalling pathway, NO binds to the heme group of soluble guanylate cyclase (sGC), a member of the guanylate cyclase family with a CYC domain (PF00211), to increase cGMP. However, no NO targets were found in cells that highly express sGC within the cPRC circuit. Recently, Moroz and others reported that GC with NIT (nitrite/nitrate sensing) domains (PF08376) (NIT-GC), identified as a sensor for nitrate and nitrite, the oxides of NO, is relatively widely conserved in metazoans (Moroz et al., 2020). To search for NIT-GCs in Platynereis, genes with the CYC domains were searched using the transcriptome and genome sequences of 45 metazoan species and choanoflagellate. By sequence-similarity searches, multiple sequence alignments and cluster analyses, a cluster of membrane-bound (guanylate cyclase-coupled receptor) type was found and a phylogenetic tree was constructed with soluble type (sGC) as an outgroup. As in the phylogenetic analysis of Moroz et al, we found a group of GCs with NIT domains that included many metazoans such as placozoa, cnidaria, ecdysozoa, echinoderms and lophotrochozoa, and in Platynereis 15 NIT-GCs were found (Fig. 3A). To find the candidate NIT-GCs involved in the cPRC circuit among the 15 NIT-GCs, we used previously published spatially mapped single-cell transcriptome data [@Achim2015][@williams2017] and Two NIT-GCs were found to be highly expressed in the cPRC, and we analysed their expression using in situ HCR as NIT-GC1 and NIT-GC2. As a result, strong signals for both were observed specifically in the vicinity of the cPRCs in 2-day-old larvae (Fig. 3B). Both c-opsin1 and MLD pNP (pedal peptide neuropeptide precursor 2), already identified as cPRC markers, and NIT-GC1 and NIT-GC2 were found to be co-expressed. They were co-expressed c-opsin1 and MLD pNP (pedal peptide neuropeptide precursor 2), already identified as cPRC markers (Fig. 3C) (Arendt et al., 2004; Williams et al., 2017). To analyse localisations at the protein level, immunostaining was performed using polyclonal antibodies against NIT-GC1 and NIT-GC2. Interestingly, NIT-GC1 was found to localise to the neuronal axon and NIT-GC2 to the ciliary region of the cPRC (Fig. 3D and E). Furthermore, co-immunostaining with NOS antibodies revealed that NOS and NIT-GC1 were localised in close proximity in the neurosecretory plexus (Fig. 3F). These results suggest that NIT-GC1 may function as a major target for NO.

>>>>

---
title: Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance
  in zooplankton
author: "Kei Jokura, Nobuo Ueda, Martin Gühmann, Gáspár Jékely"
date: '*Living Systems Institute, University of Exeter, Stocker road, Exeter, EX4
  4QD, UK*'
output:
  html_document:
    theme: lumen
  pdf_document:
    latex_engine: lualatex
    template: default
  word_document: default
geometry: left=0.8in,right=0.8in,top=0.8in,bottom=0.5in
mainfont: Arial
monofont: Times New Roman
documentclass: article
classoption: onecolumn
linestretch: 1.2
fontsize: 10pt
bibliography: references.bib
csl: elife.csl
---

@@

Nitric oxide (NO) generated by nitric-oxide synthase (NOS) is a key regulator of animal physiology. In nervous systems, NO can modulate circuits, but its effects on neural dynamics and behaviour remain unclear. Here we show how NO signalling shapes circuit dynamics to orchestrate light-avoidance behaviour. We studied UV-avoidance mediated by brain ciliary photoreceptor cells (cPRCs) in larval *Platynereis dumerilii*, a marine annelid. We found NOS specifically expressed in interneurons (INNOS) postsynaptic to cPRCs. Stimulation of cPRCs by violet light leads to cPRC inhibition but concomitant INNOS activation and NO production. NO feeds back to cPRCs and triggers delayed cPRC activation through an unconventional guanylyl cyclase. This results in the activation of a second population of postsynaptic interneurons (INRGW). In *NOS* knockout larvae, NO feedback and INRGW activation do not occur and avoidance behaviour is defective. Our results reveal how NO-mediated neuroendocrine signalling gates a synaptic circuit to initiate light-avoidance behaviour.

@@

The ability to perceive changes in the environment and adapt behaviour accordingly is essential for the survival of all living organisms. Avoidance from UV radiation, which damages DNA essential for survival, is proposed to be one of the behaviours that existed before the abundant development of Earth's oxygen to block UV radiation in the atmosphere (Pittendrigh). In the ocean, almost all zooplankton have a pronounced rhythmic behaviour called Diel vertical migration (DVM), which consists of upwards towards the surface at dusk and descend back to deeper water before dawn. Avoidance behaviour to escape harmful UV radiation during the day is closely linked to the control of vertical migration in the ocean. In metamorphosis, a life history strategy widely found in metazoans, downward swimming towards the seafloor is known to be associated with larval settlement and the initiation of metamorphosis, a dramatic body change from the larval stage, and UV-avoidance behaviour allows dispersal and the acquisition of different ecological niches. However, little is known about the molecular neural circuit mechanisms in UV avoidance behaviour in many animals, including zooplankton. In neural circuits in general light response, the activation of various receptors by excitatory and inhibitory neurotransmitters such as glutamate, GABA and dopamine is known in various species. In addition to this, it has long been known that in nitric oxide (NO), as a neuromodulatory molecule, it is also important as a modulator of visual activity. NO is a small free-radical molecule with diffusible and changeable characteristics, and thus works as one of the neural signalling molecules. The function of NO in the neural circuitry of vision has been well studied in vertebrates. In the mammalian retina, NO is an essential signalling molecule in all retinal cells. NO produced by nitric oxide synthase (NOS) in amacrine cells, one of the retinal cells, activates soluble guanylate cyclase (sGC) expressed in cone bipolar cells, which produces cGMP and regulates downstream signalling. On the other hand, the role of NO in zooplankton has been reported in many animals, including sponges, cnidarians, molluscs, arthropods, annelids, echinoderms and ascidians, as a molecule that controls settlement and metamorphosis <!---# I think metamorhosis is less relevant for the paper --> initiation. We therefore hypothesise that NO-mediated neural signalling may be involved in the photoreceptor circuits that control UV avoidance behaviour. To test this hypothesis we used the annelid *Platynereis dumerilii*, which conserved vertebrate-type ciliary photoreceptors (cPRCs) [@Arendt2004]. The cPRC expresses c-opsin1, which absorbs UV radiation, is involved in melatonin production and regulates circadian swimming behaviour [@Tosches2014; @tsukamoto2017]. Furthermore, we previously reported abnormal UV avoidance behaviour in c-opsin1 knockout larvae [@Veraszto2018]. The cPRCs express the UV-absorbing c-opsin1 and mediate UV avoidance in the larvae [@Veraszto2018]. In adult worms, UV light and c-opsin1 modulate locomotion and maturation [@Veedin2021]. These suggest that Platynereis larvae are a highly suitable model for elucidating the neural circuit mechanisms of NO-induced UV avoidance behaviour, and the accessibility of genome editing, behavioural assays, calcium imaging and whole-body connectomics could be very useful for study at the molecular level[@ozpolat2021].

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data. Phylogenetic analysis of NOS proteins indicate that the *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences <!---# need to add a NOS tree as supplement -->. To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis. In two-day-old larvae, we detected *NOS* expression in two cells in the apical organ region (Fig. 1A) <!---# 3d HCR and transgene? -->. The same two cells cells were labelled with a reporter construct driving palmitoylated-tdTomato by a 12 kbp upstream genomic region of the *NOS* gene. This reporter also revealed the axonal projections of the NOS-expressing neurons. In three-day-old larvae...  <!---# need 3d HCR and transgene data as well -->

The position and morphology of the four *NOS*-expressing cells allowed us to identify the same cells in the three-day-old whole-body volume EM data [@williams2017; @veraszto2020] as four interneurons (INNOS)(Figure 1D-E). The INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW interneurons, which are also cPRC targets (Fig. 1F).

@@

**Nitric oxide signalling mediates UV-avoidance behaviour** The specific expression of NOS in interneurons directly postsynaptic to the cPRCs suggests that NO signalling is involved in the UV response. To test this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method. We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both leading to an early stop codon and thus likely representing null alleles (Figure 3A). We could establish a homozygous line for both mutations indicating that NOS is not an essential gene. To quantify UV avoidance, we recorded the trajectories of freely swimming larvae in vertical columns, illuminated laterally from two opposite sides with xy nm violet light. As previously shown, wild-type larvae swim downward within 30 sec following UV/violet light stimulation. In contrast, homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response (Figure 3B). This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae and reveals a requirement for NOS in UV-avoidance behaviour.

![**Figure 2. Larvae with inhibited NOS have shorter ciliary arrest times and a delayed response to UV-avoidance.** (A) Percentage of time in which cilia are closed in DMSO-treated larvae (control) vs. larvae exposed to NOS inhibitor. (B) Top: The domain structure of NOS. Bottom: Close-up region showing the genomic locus of NOS gene and the wild-type sequence (WT) targeted by CRISPR/Cas9. The generated mutants (NOSΔ11, NOSΔ23) are also shown. Pink indicates target sites. Yellow shows PAM sequences, red shows stop codons. (C and D) Vertical swimming in wild type and mutant larvae at 2-day-old (C) and 3-day-old (D) stimulated with blue (480 nm) and UV (395 nm) light from above. the data are shown in 30 s bins.](../figures/Fig2.png){alt="Figure 2. Larvae with inhibited NOS have shorter ciliary arrest times and a delayed response to UV-avoidance. (A) Percentage of time in which cilia are closed in DMSO-treated larvae (control) vs. larvae exposed to NOS inhibitor. (B) Top: The domain structure of NOS. Bottom: Close-up region showing the genomic locus of NOS gene and the wild-type sequence (WT) targeted by CRISPR/Cas9. The generated mutants (NOSΔ11, NOSΔ23) are also shown. Pink indicates target sites. Yellow shows PAM sequences, red shows stop codons. (C and D) Vertical swimming in wild type and mutant larvae at 2-day-old (C) and 3-day-old (D) stimulated with blue (480 nm) and UV (395 nm) light from above. the data are shown in 30 s bins." width="100%"}

@@

To test if NO is produced during UV/violet responses, we injected zygotes with the fluorescent NO-reporter DAF-FM and imaged two-day-old larvae during the local stimulation of the cPRC cilia by 405 nm light. Following light stimulation, we detected an increase in DAF-FM fluorescence in the region of the INNOS cells. (#Kei - are there time course data that could be shown?). We next analysed the dynamics of the cPRC circuit following light stimulation. As we have shown previously [@Veraszto2018], a 30-sec illumination of the cPRC ciliary region lead to a decrease in cPRC calcium levels, likely due to the hyperpolarisation of the cPRCs. After 30 min, calcium levels in cPRCs were raising again, reaching higher levels than at the start of the stimulus - a response that may involve depolarisation. This activation phase occurs after the end of the stimulation and may be due to a slower neuroendocrine feedback [@Veraszto2018]. To determine whether NO mediates such a feedback, we measured calcium signals in *NOS*-mutant larvae. While we detected the inhibition phase in cPRC in *NOS*-mutants, this was not followed by activation and calcium levels remained low. These results revealed an essential requirement for NOS signalling in the second phase of the cPRC response.

We next quantified the responses of the INNOS and INRGW interneurons following a UV/violet stimulus to the cPRCs. To unambiguously identify the cells from which we recorded calcium signals, we developed an on-slide immunostaining method. We used the cell-specific neuropeptide markers anti-RYamide (INNOS) and anti-RGWa (INRGW) [@Conzelmann2011] to mark the interneurons. Based on the position of the nuclei, we could correlate live and fixed samples at a single-cell precision (Figure 2 - figure supplement ). In both wild type and *NOS*-mutant larvae, INNOS cells showed increase calcium signals during the inhibition phase of the cPRCs. In contrast, calcium signals increased in INRGW cells in wild-type larvae during the second activation phase of cPRCs, a response that was lacking in *NOS*-mutants. This suggests that there is no circuit output to the Ser-h1 cells and the ciliary band in *NOS*-mutants, consistent with the observed behavioural defects.

@@

NO generally acts via cyclic guanosine monophosphate (cGMP) as a second messenger. In this signalling pathway, NO binds to the heme group of soluble guanylate cyclase (sGC), a member of the guanylate cyclase family with a CYC domain (PF00211), to increase cGMP. However, no NO targets were found in cells that highly express sGC within the cPRC circuit. Recently, Moroz and others reported that GC with NIT (nitrite/nitrate sensing) domains (PF08376) (NIT-GC), identified as a sensor for nitrate and nitrite, the oxides of NO, is relatively widely conserved in metazoans (Moroz et al., 2020). To search for NIT-GCs in Platynereis, genes with the CYC domains were searched using the transcriptome and genome sequences of 45 metazoan species and choanoflagellate. By sequence-similarity searches, multiple sequence alignments and cluster analyses, a cluster of membrane-bound (guanylate cyclase-coupled receptor) type was found and a phylogenetic tree was constructed with soluble type (sGC) as an outgroup. As in the phylogenetic analysis of Moroz et al, we found a group of GCs with NIT domains that included many metazoans such as placozoa, cnidaria, ecdysozoa, echinoderms and lophotrochozoa, and in Platynereis 15 NIT-GCs were found (Fig. 3A). To find the candidate NIT-GCs involved in the cPRC circuit among the 15 NIT-GCs, we used previously published spatially mapped single-cell transcriptome data [@Achim2015][@williams2017] and Two NIT-GCs were found to be highly expressed in the cPRC, and we analysed their expression using in situ HCR as NIT-GC1 and NIT-GC2. As a result, strong signals for both were observed specifically in the vicinity of the cPRCs in 2-day-old larvae (Fig. 3B). Both c-opsin1 and MLD pNP (pedal peptide neuropeptide precursor 2), already identified as cPRC markers, and NIT-GC1 and NIT-GC2 were found to be co-expressed. They were co-expressed c-opsin1 and MLD pNP (pedal peptide neuropeptide precursor 2), already identified as cPRC markers (Fig. 3C) (Arendt et al., 2004; Williams et al., 2017). To analyse localisations at the protein level, immunostaining was performed using polyclonal antibodies against NIT-GC1 and NIT-GC2. Interestingly, NIT-GC1 was found to localise to the neuronal axon and NIT-GC2 to the ciliary region of the cPRC (Fig. 3D and E). Furthermore, co-immunostaining with NOS antibodies revealed that NOS and NIT-GC1 were localised in close proximity in the neurosecretory plexus (Fig. 3F). These results suggest that NIT-GC1 may function as a major target for NO.
